---
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(RColorBrewer)
library(ggExtra)
library(ggplot2)
library(ggthemes)
library(gridExtra)
library(hrbrthemes)
library(patchwork)
library(PerformanceAnalytics)
library(tidyverse)
library(vioplot)
```

# Car information dataset analysis. 
The dataset contains 399 rows of 9 features, which contains some general properties 
of cars. These 9 features are the following:

1. Name: Unique identifier for each automobile.
2. MPG: Fuel efficiency measured in miles per gallon.
3. Cylinders: Number of cylinders in the engine.
4. Displacement: Engine displacement, indicating its size or capacity.
5. Horsepower: Power output of the engine.
6. Weight: Weight of the automobile in pounds.
7. Acceleration: Capability to increase speed, measured in seconds to 60 miles/hour.
8. Model Year: Year of manufacture for the automobile model.
9. Origin: Country or region of origin for each automobile.

The dataset can be found via this 
[**link**](https://www.kaggle.com/datasets/tawfikelmetwally/automobile-dataset)

## Data exploration
In the output below, the first 5 lines of the data can be seen, and the overall structure. 
```{r}
# setwd("/media/sf_SF/Fedora/R_course/Assignment/")
car_data <- read.csv("Automobile.csv")
head(car_data)
str(car_data)
```

Let's check if there are empty values, while omitting these directly. 
```{r}
missing_values <- which(is.na(car_data), arr.ind = TRUE)
print(missing_values)
print(paste("Column 5 is",colnames(car_data)[5]))
car_data <- na.omit(car_data)
```

There are six missing values in the "horsepower" column.

The dataset contains four categorical variables (name, model_year, cylinders, and origin) and six numerical variables (mpg, displacement, horsepower, weight, and acceleration).

IMPORTANT NOTE: The number of cylinders falls under categorical variables, as it divides the cars into categorical groups based on their engine. Additionally, the miles per gallon will be disregarded, as the mpg values will be converted into liters per 100 kilometers (L/100km) and placed within a new column.

```{r}
car_data$model_year <- as.character(car_data$model_year)
car_data$model_year <- paste0("19",car_data$model_year)
car_data$cylinders <- as.character(car_data$cylinders)
car_data$L_100km <- 235.215 / car_data$mpg
cat_var <- c("name", "brand", "model_year", "origin", "cylinders")
num_var <- c("L_100km", "displacement", "horsepower", "weight", "acceleration")
```

Let's take a look at the frequencies of each categorical variable. Because of 
the huge amount of unique car models, no representative barplot can be generated.
The complete dataset contains 37 car brands

```{r}
car_data$brand <- sapply(strsplit(car_data$name, " "), `[`, 1)

paste0("Unique car models: ",length(unique(car_data$name)),', Unique car brands: ', length(unique(car_data$brand)))
```
```{r, warning=FALSE}
par(mfrow = c(1,2))
bp <- barplot(table(car_data$origin),
        main = "Frequency of each origin",
        ylim = c(0,max(table(car_data$origin))+50),
        col = brewer.pal(3, "Paired"))
text(x=bp, y=table(car_data$origin),label=table(car_data$origin),pos=3)
bp <- barplot(table(car_data$cylinders),
        main = "Frequency of numb of cylinders",
        xlab = "Amount of cylinders",
        ylim = c(0,max(table(car_data$cylinders)+20)),
        col = brewer.pal(3, "Paired"))
text(x=bp, y=table(car_data$cylinders),label=table(car_data$cylinders),pos=3)
par(mfrow = c(1,1))
bp <- barplot(table(car_data$model_year),
        main = "Frequency of each model year",
        ylim = c(0,45),
        col = brewer.pal(12, "Set1"),
        las = 2)
text(x=bp, y=table(car_data$model_year),label=table(car_data$model_year),pos=3)
```

The plots above show that the model years are fairly normaly distributed. But on the other hand, the origin and amount of cylinders of the cars are not equally distributed. 
Let's take a look at the density and distributions of the numerical variables now. 

```{r, warning=FALSE, fig.width=8.27, fig.height=11.69}
plot_list <- list() # Making a list to arrange the plots nicely later
for (var in num_var) {
  p <- ggplot(car_data, aes(x = !!sym(var))) +
    geom_density(fill = "skyblue", color = "black") +
    labs(title = paste("Densityplot of", var)) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  plot_list[[length(plot_list) + 1]] <- p
}
grid.arrange(grobs = plot_list, ncol = 2, nrow = 3) # arranging the plots in one figure
```

The acceleration variable has a normal distribution. The 4 other numerical variables all have a right skewness.

## Question 1
With this dataset, some genereal research questions can be asked. Firstly, do the amount of cylinders have a correlation with the numerical variables? To start, let's make some boxplots of the numerical variables, grouped by the amount of cylinders of the engine.  

```{r, fig.width=8.27, fig.height=5.5}
for (i in num_var) {
  # Making a table with the frequencies of each number of cylinders
  measurements <- car_data %>%
  group_by(cylinders) %>%
  summarise(count = sum(!is.na(!!sym(i))))
  
  p <- ggplot(car_data, aes(x = as.factor(cylinders), y = !!sym(i))) +
    geom_boxplot(fill = "skyblue", color = "black") +
    # Adding the frequency for the amount of cylinders
    geom_text(data = measurements, aes(label = paste(count, "Measurements"), 
                                       x = as.factor(cylinders), 
                                       y = max(car_data[[i]], na.rm = TRUE)),
              vjust = -1, size = 3) +
    labs(title = paste("Cylinders ~", i), x = "Amount of cylinders") +
    theme_minimal() + 
     theme(plot.title = element_text(hjust = 0.5))
  plot_list[[length(plot_list) + 1]] <- p
  print(p)
}
```

As the plots above depict, the median of fuel consumption (excluding the three cylinders), displacement, horsepower and weight increase with the number of cylinders for all variables, except for the acceleration of the cars. 

Now, what about the density of each variable plotted over each other? Can we see the same pattern as clearly here?

```{r, warning=FALSE, fig.width=8.27, fig.height=5.5}
for (i in num_var){
  p <- ggplot(data = car_data, aes(x = car_data[,i], fill = as.factor(cylinders))) +
    geom_density(alpha = 0.7) +
    labs(title = paste("Density of",i,"grouped by cylinders"), 
         x = i,
         fill = "Cylinders") +
    theme_base() +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "top")
  print(p)
}
```

Here again, we observe a similar trend, with the exception of the acceleration variable. The density plots shift to the right as the number of cylinders increases. However, not every density plot is visually pleasing due to the limited sample size for 3 and 5 cylinders. 
Next, let's further investigate some correlatations, but now via a correlationplot of the numerical variables only.

```{r, warning=FALSE}
chart.Correlation(car_data[num_var], 
                  method = "pearson",
                  histogram = TRUE, 
                  pch = 16)
```

The correlation plot provides a comprehensive overview of the numerical variables that are strongly related through scatterplots, densigrams, and correlation coefficients. Now, let's narrow down our focus to the numerical variables that have an absolute correlation coefficient greater than 0.80.

```{r, message=FALSE, warning=FALSE, fig.width=8.27, fig.height=5.5}
# Note that the code for only one plot is shown. 
p <- ggplot(car_data ,aes_string(x = "L_100km", 
                                 y = "displacement", 
                                 color = "cylinders")) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) + 
  theme(legend.position = "right") +  
  labs(color = "Cylinders") 
      
p2 <- ggMarginal(p, type = "densigram")
print(p2)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=8.27, fig.height=5.5}
p <- ggplot(car_data, aes_string(x = "L_100km", 
                                 y = "horsepower", 
                                 color = "cylinders")) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) + 
  theme(legend.position = "right") +  
  labs(color = "Cylinders") 
      
p2 <- ggMarginal(p, type = "densigram")
print(p2)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=8.27, fig.height=5.5}
p <- ggplot(car_data ,aes_string(x = "L_100km", 
                                 y = "weight", 
                                 color = "cylinders")) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) + 
  theme(legend.position = "right") +  
  labs(color = "Cylinders") 
      
p2 <- ggMarginal(p, type = "densigram")
print(p2)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=8.27, fig.height=5.5}
p <- ggplot(car_data ,aes_string(x = "displacement", 
                                 y = "horsepower", 
                                 color = "cylinders")) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) + 
  theme(legend.position = "right") +  
  labs(color = "Cylinders") 
      
p2 <- ggMarginal(p, type = "densigram")
print(p2)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=8.27, fig.height=5.5}
p <- ggplot(car_data ,aes_string(x = "displacement", 
                                 y = "weight", 
                                 color = "cylinders")) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) + 
  theme(legend.position = "right") +  
  labs(color = "Cylinders") 
      
p2 <- ggMarginal(p, type = "densigram")
print(p2)
```
```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=8.27, fig.height=5.5}
p <- ggplot(car_data ,aes_string(x = "horsepower", 
                                 y = "weight", 
                                 color = "cylinders")) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) + 
  theme(legend.position = "right") +  
  labs(color = "Cylinders") 
      
p2 <- ggMarginal(p, type = "densigram")
print(p2)
```

The red line in the scatter plot represents the fitting of a linear model to the data. The width of the red area surrounding the red line indicates the variance in the data for that particular point. A wider red area suggests higher variability in the data at that point.


Based on the analysis of the previous plots, it can be concluded that the number of cylinders has a positive impact on the numerical variables. Additionally, the strong correlations observed in the last six plots further support this conclusion.

## Question 2
Another research question to explore is whether there are any differences in the numerical variables among the origins of the cars. To investigate this, a violin plot was created.

```{r, fig.width=8.27, fig.height=11.69}
par(mfrow = c(3,2))
for (i in num_var) {
  p <- vioplot(car_data[,i] ~ car_data$origin,
               col = brewer.pal(3, "Paired"),
               xlab = "Region",
               ylab = paste(i),
               main = paste("Region ~",i))
  p
}
```
The violin plots above illustrate that cars in America tend to have higher fuel consumption, larger engines, more horsepower, and greater weight compared to those in Europe and Japan. However, they have similar acceleration characteristics. It's important to note that the dataset contains a higher proportion of cars from the USA, resulting in a larger variance for each numerical variable. This observation is also evident in the violin plots, where the plots for the USA have longer tails compared to the other two regions.

Considering our previous findings, it's plausible that there are more 8-cylinder cars in the USA. Let's explore this hypothesis further with a stacked barplot.

```{r, warning=FALSE}
barplot(table(car_data[,c("origin", "cylinders")]),
        main = "Barplot of the frequency of region per amount of cylinders",
        xlab = "Amount of cylinders",
        ylim = c(0,200), 
        ylab = "Frequency",
        col = brewer.pal(3, "Paired"))
legend("topright", legend = rownames(table(car_data$origin)), fill = brewer.pal(3, "Paired"))
```
As depicted above, all 8-cylinder and almost all 6-cylinder cars originate from America, indicating heavier and more fuel-consuming engines.

## Conclusion via PCA
Lastly, let's investigate whether there is clustering based on the origin of the car or/and  based on the number of cylinders, confirming the findings from research questions 1 and 2. To conduct this analysis, a principal component analysis (PCA) was performed.
```{r}
pca_result <- prcomp(na.omit(car_data[num_var]), scale. = TRUE, center = TRUE)
pca_df <- as.data.frame(pca_result$x)
pca_df <- cbind(pca_df, origin = car_data$origin)

# Create the PCA plot with origin coloring
ggplot(pca_df, aes(x = PC1, y = PC2, color = origin)) +
  geom_point(size = 3) +
  labs(title = "PCA Plot of Numerical variables, colored by origin",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()

pca_df <- cbind(pca_df, cylinders = car_data$cylinders)
# Create the PCA plot with cylinder coloring
ggplot(pca_df, aes(x = PC1, y = PC2, color = cylinders)) +
  geom_point(size = 3) +
  labs(title = "PCA Plot of Numerical variables, colored by number of cylinders",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()
```

In the PCA plot, two distinct clusters are noticeable—one on the left and one on the right. Upon examining the plot where the number of cylinders is color-coded, a clear separation can be observed on the left side, corresponding to the 8-cylinder category, which is distinct from the rest of the dataset. It's noteworthy that all data points in this cluster belong to the USA region, as evident from the very first PCA plot. Overall, it appears that the number of cylinders serves as a clearer clustering factor compared to the region from which the car originated.



